---
title: "Getting Started with spatialbp"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with spatialbp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The `spatialbp` package is a high-performance R-Rust framework designed to identify **2D continuous spatial boundaries** (piecewise regression changepoints) in histology, multiplexed immunofluorescence (mIF), and spatial transcriptomics datasets.

Rather than drawing rigid, manual Regions of Interest (ROIs), `spatialbp` uses a high-performance **Markov Chain Monte Carlo (MCMC) engine** written in Rust to dynamically discover continuous shared spatial boundaries based on the spatial gradients of your biological markers.

---

## 1. Installation & Setup

To load the package in your active R session, run:

```{r}
library(spatialbp)
```

*Note for Windows users:* The package relies on a custom-designed Rtools/GCC linker built directly into the package's compilation pipeline, ensuring zero-dependency compilation.

---

## 2. Quick Start: Fitting a Single-Channel Boundary

For biologists and lab scientists, `spatialbp` provides a simplified high-level wrapper function called `fit_spatial_boundary()`. This function automates all coordinates projection and MCMC step-size calibration.

Let's simulate a standard 2D cortical gliosis tissue where astrocytes (GFAP) are highly reactive in the center:

```{r, fig.width = 6, fig.height = 5}
library(spatialbp)

# Simulate cortical tissue grid (100x100 resolution)
df <- simulate_cortical_tissue(n_grid = 100, noise = 0.2)

# Fit the boundary in 1 simple line!
fit_astro <- fit_spatial_boundary(
  df = df, 
  y1 = "Astrocytes", 
  latent_type = "radial", 
  iterations = 3000
)

# Plot the beautiful fitted boundary line
plot(fit_astro)
```

The S3 plotting method automatically renders a gorgeous, high-contrast overlay of the estimated spatial boundary on top of your tissue intensity coordinates!

---

## 3. Joint Spatial Modeling: Affinity vs. Repulsion

In spatial biology, markers often interact. The package supports **Joint Spatial Modeling** by tying the boundary parameters ($c$) across multiple channels:

### A. Spatial Affinity (Co-localization)
Two proteins (e.g. Astrocytes and Microglia) share the **same boundary** and exist on the **same side** of that boundary:
* $Y_1 = \beta_1 \max(0, Z - c)$
* $Y_2 = \beta_2 \max(0, Z - c)$

### B. Spatial Repulsion (Exclusion/Segregation)
Two markers (e.g. Microglia and Perineuronal Networks) share a physical boundary but are active on **opposite sides** of it:
* $Y_1 = \beta_1 \max(0, Z - c)$ (Inside the boundary)
* $Y_2 = \beta_2 \max(0, c - Z)$ (Outside the boundary)

`fit_spatial_boundary()` automatically detects if your channels co-localize or segregate using their correlation, and runs the appropriate joint R-Rust MCMC fit:

```{r, fig.width = 6, fig.height = 5}
# Fit a joint repulsion model for Microglia and PNNs
fit_joint <- fit_spatial_boundary(
  df = df,
  y1 = "Microglia",
  y2 = "PNNs",
  latent_type = "radial",
  iterations = 3000
)

# Plot the joint boundary
plot(fit_joint)
```

---

## 4. Advanced: Automatic Relationship Discovery

If you have many channels, you can automatically discover their spatial interactions without manually running pairs. The `discover_spatial_relationships()` function runs combinatorial pairwise joint models and outputs a **Bayesian Spatial Network**:

```{r}
# Define a list of your stains
stains_list <- list(
  Astrocytes = df$Astrocytes,
  Microglia = df$Microglia,
  PNNs = df$PNNs
)

# Run discovery
results <- discover_spatial_relationships(
  stains = stains_list,
  latent_field = df$LatentInflame,
  iterations = 2000,
  bf_threshold = 10
)

# View the automatically classified network
print(results$relationships)
```

This returns a clear table classifying each pair as **Affinity**, **Repulsion**, or **Independent** based on their joint MCMC Bayes Factors!

---

## 5. Bayesian MCMC Diagnostics

To ensure your Bayesian models are valid and have reached stationarity, always check your MCMC trace plots:

```{r, fig.width = 6, fig.height = 6}
# Plot MCMC traces and posterior densities
plot_diagnostics(fit_astro, warmup = 1000)
```

---

## 6. Statistical Foundation (For Advanced Users)

### The Piecewise Regression Likelihood
Our framework models spatial intensity $Y$ using a continuous 1D latent coordinate $Z$ (e.g. cortical depth, radial distance from a center, or radial basis values):

$$Y_i = \text{baseline}_i + \beta \max(0, Z_i - c) + \epsilon_i \quad \text{where} \quad \epsilon_i \sim \mathcal{N}(0, \sigma^2)$$

### Laplace Priors (L1 Regularization)
To support noise-filtering and handle multiple boundaries in complex datasets, `spatialbp` includes L1 regularization (Lasso priors) on the slope parameters $\beta_k$:

$$\text{log\_likelihood} = -0.5 \cdot \text{SSE} - \sum_k \lambda_k |\beta_k|$$

This shrinks inactive boundary channels to exactly 0, filtering out noise in multi-boundary environments!

### High-Performance Rust MCMC Engine
Because evaluating high-resolution grids ($10,000+$ pixels/voxels) for thousands of MCMC steps is computationally expensive in R, our likelihood calculation and random-walk Metropolis-Hastings loop are executed inside a **highly-optimized compiled Rust backend** using `extendr`. This completes thousands of iterations in under 2 seconds!
