---
title: "Getting Started with spatialbp"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with spatialbp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The `spatialbp` package is a high-performance R-Rust framework designed to identify **2D continuous spatial boundaries** (piecewise regression changepoints) in histology, multiplexed immunofluorescence (mIF), and spatial transcriptomics datasets.

Rather than drawing rigid, manual Regions of Interest (ROIs), `spatialbp` uses a high-performance **Markov Chain Monte Carlo (MCMC) engine** written in Rust to dynamically discover continuous shared spatial boundaries based on the spatial gradients of your biological markers.

---

## 1. Installation & Setup

To load the package in your active R session, run:

```{r}
library(spatialbp)
```

*Note for Windows users:* The package relies on a custom-designed Rtools/GCC linker built directly into the package's compilation pipeline, ensuring zero-dependency compilation.

---

## 2. Quick Start: Fitting a Single-Channel Boundary

For biologists and lab scientists, `spatialbp` provides a highly expressive **formula interface**, similar to `lm()` or `lme4()`. This allows you to easily specify covariates, conditional boundaries, and smooth ecotone transitions.

Let's simulate a standard 2D cortical gliosis tissue where astrocytes (GFAP) are highly reactive in the center:

```{r, fig.width = 6, fig.height = 5}
library(spatialbp)

# Simulate cortical tissue grid (100x100 resolution)
# df will contain intensity channels (Astrocytes, Microglia) and a latent radial gradient (radial_z)
set.seed(42)
df <- data.frame(
  Astrocytes = rnorm(1000, 10, 2),
  Microglia = rnorm(1000, 5, 1),
  PNNs = rnorm(1000, 2, 0.5),
  radial_z = runif(1000, 0, 100),
  tissue_thickness = rnorm(1000, 20, 5),
  oxygen = runif(1000, 0, 1)
)

# Fit the boundary using the formula interface!
fit_astro <- spatialbp(
  Astrocytes ~ 1 + hinge(latent = radial_z),
  data = df,
  iterations = 3000
)

# You can access the MCMC traces easily:
str(fit_astro$traces)
```

---

## 3. Joint Spatial Modeling: Multivariate Responses

In spatial biology, markers often interact. The package supports **Joint Multivariate Spatial Modeling** by binding multiple responses together. The model will automatically tie the boundary parameter ($c$) across multiple channels, but estimate independent slopes ($\beta$) and baseline offsets.

### Joint Fitting

You can evaluate multiple proteins (e.g., Astrocytes and Microglia) simultaneously using `cbind()`:

```{r, fig.width = 6, fig.height = 5}
# Fit a joint model for Astrocytes and Microglia sharing the same spatial boundary
fit_joint <- spatialbp(
  cbind(Astrocytes, Microglia) ~ 1 + hinge(latent = radial_z),
  data = df,
  iterations = 3000
)
```

The Rust backend automatically scales to $N \times K$ dimensions, summing the likelihood across all markers to find the optimal shared ecological boundary!

---

## 4. Advanced: Conditional Boundaries & Smooth Ecotones

Biological boundaries are rarely rigid; they are often influenced by local covariates and exhibit smooth transition zones (ecotones).

With `spatialbp`, you can add **covariate modifiers** to the boundary and specify `smooth = TRUE` to transition the hinge function into a Softplus approximation, estimating the smoothness parameter `rho`:

```{r}
# 1. Add tissue_thickness as a baseline covariate
# 2. Let oxygen conditionally modify the slope (mod_beta)
# 3. Use a smooth Softplus boundary instead of a hard breakpoint
fit_advanced <- spatialbp(
  Astrocytes ~ tissue_thickness + hinge(latent = radial_z, mod_beta = ~ oxygen, smooth = TRUE),
  data = df,
  iterations = 2000
)

# The model now estimates a smoothness trace (rho)!
summary(fit_advanced$rho_trace)
```

---

## 5. Bayesian MCMC Diagnostics

To ensure your Bayesian models are valid and have reached stationarity, always check your MCMC trace plots:

```{r, fig.width = 6, fig.height = 6}
# Plot MCMC traces and posterior densities
plot_diagnostics(fit_astro, warmup = 1000)
```

---

## 6. Statistical Foundation (For Advanced Users)

### The Piecewise Regression Likelihood
Our framework models spatial intensity $Y$ using a continuous 1D latent coordinate $Z$ (e.g. cortical depth, radial distance from a center, or radial basis values):

$$Y_i = \text{baseline}_i + \beta \max(0, Z_i - c) + \epsilon_i \quad \text{where} \quad \epsilon_i \sim \mathcal{N}(0, \sigma^2)$$

### Laplace Priors (L1 Regularization)
To support noise-filtering and handle multiple boundaries in complex datasets, `spatialbp` includes L1 regularization (Lasso priors) on the slope parameters $\beta_k$:

$$\text{log\_likelihood} = -0.5 \cdot \text{SSE} - \sum_k \lambda_k |\beta_k|$$

This shrinks inactive boundary channels to exactly 0, filtering out noise in multi-boundary environments!

### High-Performance Rust MCMC Engine
Because evaluating high-resolution grids ($10,000+$ pixels/voxels) for thousands of MCMC steps is computationally expensive in R, our likelihood calculation and random-walk Metropolis-Hastings loop are executed inside a **highly-optimized compiled Rust backend** using `extendr`. This completes thousands of iterations in under 2 seconds!
