---
title: "Organic Topologies: Density and Distance Fields"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Organic Topologies: Density and Distance Fields}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Real-world biological and ecological communities rarely conform to simple geometric boundaries (like straight lines or perfect circles). Organisms often form scattered, highly irregular, organic patches across a landscape. 

When dealing with such chaotic spatial distributions, we can abandon geometric coordinates (X/Y) entirely. Instead, we project the boundaries into **Density Space** or **Distance Space**. 

By utilizing continuous topographical fields, `spatialbp` can effortlessly identify complex, asymmetrical boundaries wrapping around dozens of scattered communities simultaneously!

## 1. Simulating an Organic, Scattered Landscape

Let's generate a synthetic ecosystem where:
1. **Green (Host)**: Forms 5 random, scattered patches across the landscape.
2. **Red (Affinity)**: Forms scattered patches that strongly cluster *around* the Green patches.
3. **Blue (Repulsion)**: Forms scattered patches strictly in the voids *away* from Green patches.

```{r, fig.width=9, fig.height=3.5}
library(spatialbp)
library(ggplot2)
library(patchwork)

set.seed(123)

# 1. Generate Community "Seeds"
green_seeds <- data.frame(x = runif(5, 10, 90), y = runif(5, 10, 90))

# Red seeds cluster near Green (Affinity)
red_seeds <- data.frame(
  x = green_seeds$x + rnorm(5, 0, 8),
  y = green_seeds$y + rnorm(5, 0, 8)
)

# Blue seeds avoid Green (Repulsion)
blue_seeds <- data.frame(
  x = c(10, 90, 50, 20, 80),
  y = c(90, 10, 90, 20, 80)
)

# 2. Build the Spatial Raster Grid
grid_size <- 100
df <- expand.grid(x = 1:grid_size, y = 1:grid_size)

# Helper function to compute inverse-distance-weighted density from seeds
compute_colony <- function(seeds, grid_x, grid_y, scale=10) {
  dens <- rep(0, length(grid_x))
  for(i in 1:nrow(seeds)) {
    d <- sqrt((grid_x - seeds$x[i])^2 + (grid_y - seeds$y[i])^2)
    dens <- dens + exp(-d^2 / (2 * scale^2))
  }
  return(dens * 50) # Scale intensity
}

df$Green <- compute_colony(green_seeds, df$x, df$y, scale=10) + rnorm(nrow(df), 0, 2)
df$Red   <- compute_colony(red_seeds, df$x, df$y, scale=12) + rnorm(nrow(df), 0, 2)
df$Blue  <- compute_colony(blue_seeds, df$x, df$y, scale=15) + rnorm(nrow(df), 0, 2)

df$Green <- pmax(0, df$Green)
df$Red   <- pmax(0, df$Red)
df$Blue  <- pmax(0, df$Blue)

# Visualize the chaotic patches
p_green <- ggplot(df, aes(x, y, fill = Green)) + geom_raster() + scale_fill_viridis_c(option="mako") + theme_void() + labs(title="Green (Host)")
p_red <- ggplot(df, aes(x, y, fill = Red)) + geom_raster() + scale_fill_viridis_c(option="inferno") + theme_void() + labs(title="Red (Affinity)")
p_blue <- ggplot(df, aes(x, y, fill = Blue)) + geom_raster() + scale_fill_viridis_c(option="plasma") + theme_void() + labs(title="Blue (Repulsion)")

p_green | p_red | p_blue
```

## 2. Using Distance Space

Instead of manually drawing regions of interest, we can compute the continuous **Nearest-Neighbor Distance Field** relative to the `Green` host. We use the built-in `compute_latent_distance` helper.

```{r, fig.width=8, fig.height=4}
# Extract the coordinates of high-density Green regions to act as our "Host Sources"
green_sources <- subset(df, Green > 25)

# Compute the distance from every pixel to the nearest Green source
df$Dist_to_Green <- compute_latent_distance(df$x, df$y, green_sources$x, green_sources$y)

# Fit the multi-species interaction against Distance Space
fit_distance <- spatialbp(
  cbind(Red, Blue) ~ Dist_to_Green + hinge(latent = Dist_to_Green),
  data = df,
  iterations = 2000, warmup = 1000
)

# Plot the results
plot(fit_distance, facet_by="none")
```

The model automatically discovered the exact distance radius surrounding **every single scattered Green colony simultaneously**.

### The `confint()` Method
You can extract Highest Density Intervals (HDIs) for the estimated boundary parameter `c` using the standard `confint()` S3 method:

```{r}
confint(fit_distance, parm = "cp", level = 0.95)
```


## 3. Using Density Space (Kernel Trick)

Alternatively, we can abandon physical distance entirely and map the relationships purely in **Density Space**. We define the latent field as the Kernel Density Estimate (KDE) of the `Green` communities using `compute_latent_density`.

```{r, fig.width=8, fig.height=4}
# Compute the Kernel Density Field of the Host
df$KDE_Green <- compute_latent_density(df$x, df$y, green_sources$x, green_sources$y)

# Fit the multi-species interaction against Density Space
fit_density <- spatialbp(
  cbind(Red, Blue) ~ KDE_Green + hinge(latent = KDE_Green),
  data = df,
  iterations = 2000, warmup = 1000
)

# Extract 95% Confidence Intervals for the Density Threshold
confint(fit_density, parm = "cp", level = 0.95)
```

By substituting geometric constraints with topographical fields (Distance or Density), `spatialbp` translates chaotic, multi-patch landscapes into elegant mathematical rules!
