---
title: "Spatial Ecotones & Mycorrhizal Fungi Symbiosis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatial Ecotones & Mycorrhizal Fungi Symbiosis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In spatial ecology, many organisms are strictly bounded by environmental factors or symbiotic relationships, creating sharp transition zones called **ecotones**. 

In this vignette, we demonstrate how `spatialbp` can identify the precise physical extent of a symbiotic network. We will use the classic **`sporophores`** dataset from the `spatstat` package, which records the exact spatial locations of fruiting bodies of different mycorrhizal fungi species surrounding a single, central young birch tree.

Because these fungi are strictly symbiotic and rely on the tree's root system for survival, they exhibit a massive density spike near the tree and drop off rapidly to zero at the edge of the root zone, forming a perfect test case for continuous Bayesian piecewise regression.

---

## 1. Preparing the Fungal Density Grids

First, we will load the point pattern, extract our two primary fungal species (*Laccaria laccata* and *Hebeloma spp.*), and use Kernel Density Estimation (KDE) to convert their discrete locations into a continuous ecological density surface grid.

```{r, message=FALSE, warning=FALSE}
library(spatialbp)
library(ggplot2)
library(spatstat.data)
library(spatstat.explore)

# Load the classic sporophores spatial point pattern
data(sporophores)

# Split the multitype point pattern into separate species
fungi_split <- split(sporophores)
laccaria <- fungi_split[["L laccata"]]
hebeloma <- fungi_split[["Hebloma spp"]]

# Convert discrete locations into continuous spatial density surfaces (KDE)
# We use a 10m bandwidth to smooth the fruiting bodies into a cohesive root network
dens_lac <- as.data.frame(density(laccaria, sigma = 10, eps = 1.0))
dens_heb <- as.data.frame(density(hebeloma, sigma = 10, eps = 1.0))

# Merge into a single grid
df <- merge(dens_lac, dens_heb, by = c("x", "y"))
names(df) <- c("x", "y", "Laccaria_Density", "Hebeloma_Density")

# Calculate the continuous ecological gradient: Distance to the central Birch Tree (0,0)
df$Distance_To_Tree <- sqrt(df$x^2 + df$y^2)

# Constrain the data strictly to the actual circular study window (radius = 50m)
df <- df[df$Distance_To_Tree <= 50, ]

head(df)
```

---

## 2. Visualizing the Symbiotic Network

Before running the mathematical boundary estimation, let's look at the raw ecological data. The birch tree sits exactly at coordinate `(0,0)`. 

```{r, fig.width = 6, fig.height = 5}
# Visualizing the raw density of Laccaria laccata
p_raw <- ggplot(df, aes(x = x, y = y, fill = Laccaria_Density)) +
  geom_raster() +
  scale_fill_viridis_c(option = "magma", name = "Density") +
  annotate("point", x = 0, y = 0, color = "cyan", size = 4, shape = 17) +
  annotate("text", x = 0, y = -5, label = "Birch Tree", color = "cyan", fontface = "bold") +
  coord_fixed() +
  labs(title = "Laccaria laccata Spatial Density", subtitle = "Symbiotic root network surrounding the host birch tree") +
  theme_minimal()

p_raw
```

We can clearly see the symbiotic density spike near the host tree and an ecotone drop-off as we move outwards!

---

## 3. Estimating the Spatial Symbiosis Boundary

We want to find the exact radial distance where the *Laccaria* fungal network abruptly ends (the edge of the functional root zone). We pass the continuous `Distance_To_Tree` column as our latent gradient into `fit_spatial_boundary()`.

```{r, fig.width = 6, fig.height = 5}
# Fit the exact spatial root-zone boundary for Laccaria
fit_lac <- spatialbp(
  Laccaria_Density ~ 1 + hinge(latent = Distance_To_Tree),
  data = df,
  iterations = 3000
)

opt_c_lac <- median(fit_lac$traces$cp)
cat("Estimated Laccaria boundary radius:", round(opt_c_lac, 2), "meters from the tree.\n")
```

The MCMC sampler mathematically confirms the visual drop-off, identifying the strict symbiotic boundary at a specific radius surrounding the tree.

---

## 4. Modeling Species Interactions: Joint Multivariate Boundaries

Now, we introduce the second species: *Hebeloma spp.* Do these two fungal species collaborate and occupy the exact same root zone boundary? We can evaluate this jointly!

We pass both species into `spatialbp()` simultaneously using `cbind()`. The Rust engine automatically scales to multi-dimensional responses, sharing the physical boundary parameter but computing unique baseline and slope estimates for each species!

```{r, fig.width = 6, fig.height = 5}
# Fit the Joint Species Interaction Model
fit_joint <- spatialbp(
  cbind(Laccaria_Density, Hebeloma_Density) ~ 1 + hinge(latent = Distance_To_Tree),
  data = df,
  iterations = 15000
)

opt_c_joint <- median(fit_joint$traces$cp)
cat("Estimated Joint boundary radius:", round(opt_c_joint, 2), "meters from the tree.\n")
```

**Interpretation:** The engine simultaneously fits a shared boundary for both *Laccaria* and *Hebeloma*, revealing the core symbiotic ecotone they both interact with!

---

## 5. Ecotone Scatter Gradient

To understand exactly how the MCMC sampler made this decision, we can plot the raw density values against the distance from the tree.

```{r, fig.width = 7, fig.height = 5}
# Scatter plot of Fungal Density vs Distance
p_grad <- ggplot(df, aes(x = Distance_To_Tree, y = Laccaria_Density, color = Hebeloma_Density)) +
  geom_point(alpha = 0.6, size = 2) +
  scale_color_viridis_c(option = "plasma") +
  geom_vline(xintercept = opt_c_joint, color = "cyan", linetype = "dashed", linewidth = 1.2) +
  labs(
    title = "Fungal Symbiosis Ecotone vs Radial Distance",
    subtitle = paste("Joint Boundary estimated at", round(opt_c_joint, 2), "meters"),
    x = "Distance from Birch Tree (meters)",
    y = "Laccaria Density"
  ) +
  theme_minimal()

p_grad
```

This plot beautifully showcases the classic "Hinge-Down" topology: density is maximal at the host root core (0-15m), decays exponentially across the active zone, and reaches a complete attenuation plateau at the cyan boundary line.

---

## 6. Rigorous Bayesian Diagnostics

For field scientists publishing in peer-reviewed journals, validating Bayesian model convergence is mandatory. We verify that our MCMC sampler explored the boundary space properly using our trace diagnostics:

```{r, fig.width = 6, fig.height = 6}
# Generate publication-ready MCMC trace plots and posterior densities
plot_diagnostics(fit_joint, warmup = 7500)
```

The flat, well-mixed "fuzzy caterpillar" trace plot for $c$ proves that the sampler converged rapidly and cleanly upon the true ecological boundary!
